---
title: "Simulating Survival Data"
description: |
  A short description on simulating survival-type data using the `sim.survdata` function
author:
  - name: WenYao, Mak
    url: https://www.linkedin.com/in/mak-wen-yao-51148154/
date: 01-12-2022
categories: 
  - survival analysis
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
    toc_depth: 6
    number_sections: true
theme: theme.css
draft: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)


library(coxed)
```

The main reference materials [(1](https://www.cambridge.org/core/journals/political-science-research-and-methods/article/abs/simulating-duration-data-for-the-cox-model/1945D7548766E76FB31C6C833976822E) and [2)](https://cran.r-project.org/web/packages/coxed/vignettes/simulating_survival_data.html) come from Jeffrey Harden and Jonathan Kropko. 

## Quick Guide 

The `sim.survdata` function simulates survival data (the duration) without specifying a specific distribution for the baseline hazard function, to match the inherent advantage of the Cox method. 

To simulate a simple survival dataset, use the `sim.survdata()` function with the `num.data.frames.` argument set to 1. 

```{r, eval=FALSE, results=FALSE, echo=TRUE}

simdata <- sim.survdata(N=1000, T=100, num.data.frames=1)

```

Other features can be changed too, including the number of observation (`N` argument), maximum survival time (`T` argument), covariates number (`xvars` argument), and proportion of right censored observations (`censor` argument). 

```{r, echo=TRUE, eval=TRUE, results=TRUE}

simdata <- sim.survdata(N=700, T=250, xvars=5, censor=.2, num.data.frames = 1)

```

To combine existing datasets with simulated duration and censor, used the `X` argument. Three variables (`postel`, `rgovm` and `pgovno`) from the included `martinvanberg` dataset were used.

```{r, echo=TRUE, }

mv.data <- dplyr::select(martinvanberg, postel, rgovm, pgovno)
simdata <- sim.survdata(T=100, X=mv.data, num.data.frames = 1)

head(simdata$data, 6)
```

