---
title: "Choosing the right statistical test"
description: |
  How to choose a statistical test for different types of clinical trial data (page under construction)
author:
  - name: WenYao, Mak
    url: https://www.linkedin.com/in/mak-wen-yao-51148154/
date: 2022-04-17
header-includes: 
  - \usepackage{wrapfig}
output:
  distill::distill_article:
    toc: true
    toc_float: true
    toc_depth: 6
    number_sections: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Motivation to write 

Using the right statistical test - and using it in the right way - is critical for clinical trials. The choice of a test depends on many factors (i.e hypothesis, study design, sample size etc) but the focus on the blog post is on selecting a statistical test. 

The blog will be a reference for my work but I hope it will be useful for anyone else who is involved in clinical research. 

The main reference materials come from the [Handbook of Biological Statistics](http://www.biostathandbook.com/index.html) by John H. McDonald and [Statistics at Square One](https://www.bmj.com/about-bmj/resources-readers/publications/statistics-square-one)

# TL;DR

For paired *dependent variable* (data not independent):

\vspace{-5truemm}

```{r paired, eval=TRUE, message=FALSE, warning=FALSE, fig.cap=" Statistical tests for paired observations", out.extra='trim={0cm 0cm 0cm 0cm},clip'}
library(DiagrammeR)
grViz("
digraph paired_test {
graph [overlap = true]
node [shape = box, 
      fontsize = 8,
      fontname = Helvetica, 
      height   = 0.4,
      color    = grey]
a[label='@@1']
b[label='@@2']
c[label='@@3']
d[label='@@4']
e[label='@@5']
f[label='@@6']
g[label='@@7']
h[label='@@8']

edge[color=grey, 
     arrowhead=none]
a->b->c
a->d->g
a->e->g
a->f->h
}
[1]: 'Paired \\n Dependent Variable'
[2]: 'Nominal'
[3]: 'McNemars test'
[4]: 'Ordinal'
[5]: 'Quantitative:\\n discrete or non-normal'
[6]: 'Quantitative:\\n normal'
[7]: 'Wilcoxon'
[8]: 'Paired t-test'
", height=300)
```

For *independent variables* that are also *continuous*: ([Ref: Statistics at Square One](https://www.bmj.com/about-bmj/resources-readers/publications/statistics-square-one))

```{r indcont, eval=TRUE, message=FALSE, warning=FALSE, fig.cap="Statistical tests for independent continous variables" }
grViz("
digraph continuous_DV {
graph [overlap = true]
node [shape = box, 
      fontsize = 8,
      fontname = Helvetica, 
      height   = 0.4,
      color    = grey]
a[label='@@1']
b[label='@@2']
c[label='@@3']
d[label='@@4']
e[label='@@5']
f[label='@@6']
g[label='@@7']

edge[color=grey, 
     arrowhead=none]
a->b [fontsize=8,label='Yes']
a->g [fontsize=8,label='No']
b->c [fontsize=8,label='Yes']
b->d [fontsize=8,label='No']
c->e [fontsize=8,label='Yes']
c->f [fontsize=8,label='No']
}
[1]: 'Data \\n is continuous?'
[2]: 'Data is \\n normally distributed?'
[3]: 'Equal variances?'
[4]: 'Mann-Whitney \\n U test'
[5]: '2-sample \\n t-test with \\n pooled variance'
[6]: '2-sample \\n t-test with \\n unequal variance'
[7]: 'Refer next\\n figure'
", height=500)
```

Lastly, for non-continuous (*discrete* or *categorical*): 

```{r discrete, eval=TRUE, message=FALSE, warning=FALSE, fig.cap="Statistical tests for independent discrete/categorical variables" }
grViz("
digraph discrete_DV {
graph [overlap = true]
node [shape = box, 
      fontsize = 8,
      fontname = Helvetica, 
      height   = 0.4,
      color    = grey]
a[label='@@1']
b[label='@@2']
c[label='@@3']
d[label='@@4']
e[label='@@5']
f[label='@@6']
g[label='@@7']
h[label='@@8']
i[label='@@9']

edge[color=grey, 
     arrowhead=none]
a->b [fontsize=8,label='No']
b->c [fontsize=8,label='No']
b->f [fontsize=8,label='Yes']
c->d [fontsize=8,label='No']
d->h [fontsize=8,label='Yes']
d->i [fontsize=8,label='No']
c->e [fontsize=8,label='Yes']
e->h [fontsize=8,label='Yes']
e->g [fontsize=8,label='No']
}
[1]: 'Data \\n is continuous?'
[2]: 'Data is \\n ordinal?'
[3]: 'More than 2 categories?'
[4]: 'All expected \\n frequencies >5?'
[5]: 'Most expected \\n frequencies >5?'
[6]: 'Chi-squared \\n test for trend'
[7]: 'Analyse with care \\n consider to pool \\n categories'
[8]: 'Chi-square \\n test'
[9]: 'Fisher exact \\n or Chi-square \\n with Yates correction'
", height=600)
```

---

# Inferential Statistics 

We perform statistical analysis to infer the results of a sample (obtained in the study) to the population. The technique is known as *inferential statistics*. 

Typically, we can either perform an *estimation* (with confidence interval) or a *hypothesis test*. 

## Independent *t* test

Also known as *unpaired two-samples t test*. It is used to **compared the mean** of two independent groups. 

In statistics, the *null ($H_{0}$)* and *alternative ($H_{1}$) hypothesis* can be defined as follow: 

```{r, message=F, warning=F}
xaringanExtra::use_panelset()
```

::::: {.panelset}

::: {.panel}

## Null Hypothesis {.panel-name}

$H_{0}$: $mean_{1}$ = $mean_{2}$

$H_{0}$: $mean_{1}$ $\leq$ $mean_{2}$

$H_{0}$: $mean_{1}$ $\geq$ $mean_{2}$
:::

::: {.panel}

## Alternative hypothesis {.panel-name}

$H_{0}$: $mean_{1}$ $\neq$ $mean_{2}$

$H_{0}$: $mean_{1}$ > $mean_{2}$

$H_{0}$: $mean_{1}$ < $mean_{2}$

:::

::: {.panel}

## Assumptions {.panel-name}

**There are four assumptions to fulfil**

1. Random samples 
2. Observations are independent 
3. Data are normally distributed (or sample size is big)
4. Equal (population) variances
:::
:::::

To check for normality, I will usually plot histogram for both groups and inspect visually. In this case, the *Heart Disease* data set from the [UCI Machine Learning Repository](https://archive.ics.uci.edu/ml/datasets/Heart+Disease). 

```{r dat, echo=FALSE}
dat <- read.csv("heart.csv")
```

