---
title: "GIS: Distribution of Pharmacy in Penang, Malaysia"
description: |
  How to use R to visualise places of interest (work in progress)
author:
  - name: WenYao, Mak
    url: https://www.linkedin.com/in/mak-wen-yao-51148154/
date: 08-16-2021
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
    toc_depth: 6
    number_sections: true
theme: theme.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Visualising Pharmacy Locations with R

*This is a side-project where new information will be continuously added* 

The motivation behind this project is that I wanted to visualise how competitive is in the retail pharmacy sector, especially in Penang. 

This project utilises open-source data published on the *Official Portal of Pharmaceutical Services Programme Register of Licenses*, available [here](https://www.pharmacy.gov.my/v2/en/information/register-licences.html). 

Disclaimer: To err is human. If you find any mistakes on this post please kindly let me know. You can reach me via: 

# Data Collection 

Information of all registered pharmacies (retail or wholesale) is manually collected from the website, filtering for License A (pharmacist license) in Penang state only. A total of 602 entries were collected. Among the important variables that I needed were (1) Name of the premise, (2) Address, and (3) Type of the business (retail, wholesale, or both).  

As the file size is quite small, the `.csv`{.R} sheet is read via the `read.csv()`{.R} wrapper function. The data frame is given a name "PG" via

```{r data, eval=TRUE, cache=FALSE, warning=FALSE, message=FALSE, echo=TRUE}
PG <- read.csv("PGpharm.csv")
```

Displaying the column names suggests that the information is sufficient for what I want to do here. (Note: the column names were written in Malay where 'Nama.Premis'='Premise Name' and 'Alamat.Premis'='Premise Address') 

```{r, eval=TRUE, cache=FALSE, warning=FALSE, message=FALSE, echo=TRUE}
colnames(PG)
```

Since the data is published by an official government portal, there should be no missing data but it doesn't hurt to check. 

```{r, eval=TRUE, cache=FALSE, warning=FALSE, message=FALSE, echo=TRUE}
which(is.na(PG$Nama.Premis))
which(is.na(PG$Alamat.Premis))
```

Once data completeness has been confirmed, I can move on to prepare the data for mapping. 

# GIS package in R 

I am mainly using the `mapview()`{.R} [package](https://r-spatial.github.io/mapview/) and `ggmap()` [package](https://www.rdocumentation.org/packages/ggmap/versions/2.6.1)

The first step is to convert the the address to latitude and longtidue. This is done by using the `geocode()` function. 

Note that geocoding requires tools that may not be free. Most notably is the [Google Maps API](https://developers.google.com/maps/gmp-get-started) that requires an account on the Google Cloud Platform in order to work. I am using a mixed of the public API by [OpenStreetMap](https://www.openstreetmap.org/#map=6/4.116/109.455) called [Nominatim](http://nominatim.org/release-docs/latest/api/Overview/) and Google Maps API. 

The following codes are used to convert addresses to latitude and longitude. I used `cascade_oder=c("osm","google")` to order first geocoding services with Nominatim followed by Google Maps service.  

```{r, eval=FALSE, cache=FALSE, warning=FALSE, message=FALSE, echo=TRUE,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
library(ggmap)

lat_longs3 <- PG %>%
  geocode(Alamat.Premis, method="cascade", cascade_order=c("osm","google"), lat=latitude, long=longitude)
```

Sometimes the geocoding process will fail to produce any coordinates and subsequent mapping cannot proceed. In this case, the coordinates will be manually identified and filled. 

# Mapping pharmacies 

As the coordinates of the pharmacies are complete now, I can proceed to map their locations for subsequent analysis. 

```{r, cache=TRUE, warning=FALSE, message=FALSE, echo=FALSE}
lat_longs3<-read.csv("penang.csv")
```

The next step is to convert the coordinates into sf objects with `st_as_sf()` command (of the `sf()` package)

```{r, eval=TRUE, cache=FALSE, warning=FALSE, message=FALSE, echo=TRUE}
library(sf)

pg_sf <- st_as_sf(lat_longs3, coords = c("longitude", "latitude"), crs=4326)
```

To map these pharmacies, the package `mapview()` will be needed. I opted to use the colour palette from the package `RColorBrewer`.

```{r, eval=TRUE, cache=FALSE, warning=FALSE, message=FALSE, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
library(mapview)
library(RColorBrewer)

pg_pharm <-mapview(pg_sf, zcol=c("Nama.Premis"), legend=FALSE, layer.name = "Retail & Wholesale Pharmacies in Penang", col.regions=brewer.pal(9, "Reds"))
```

The distribution of all retail & wholesale pharmacies in Penang state is shown below. Each colour dot represents a pharmacy. 

```{r, eval=TRUE, cache=FALSE, warning=FALSE, message=FALSE, echo=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
pg_pharm
```

```{r, eval=FALSE, cache=TRUE}

library(tm)
library(data.table)
library(stringi)
library(wordcloud)

library(ggmap)
library(tidygeocoder)
library(dplyr)
library(tibble)

library(expss)
library(lattice)
library(sp)
library(leafpop)

library(sf)
library(mapview)
library(leaflet)
library(tidyverse)
library(magrittr)
library(leafem)
library(ggplot2)
library(RColorBrewer)

lat_longs3<-read.csv("penang.csv")

pg_sf <- st_as_sf(lat_longs3, coords = c("longitude", "latitude"), crs=4326)


#Mapping 
mapviewOptions(fgb = FALSE)
my <-readRDS("gadm36_MYS_2_sf.rds", refhook = NULL)
MY <- mapview(my, legend=FALSE, layer.name="MY")

pg_pharm <-mapview(pg_sf, zcol=c("Nama.Premis"), legend=FALSE, layer.name = "Retail & Wholesale Pharmacies in Penang", col.regions=brewer.pal(9, "Reds"))

map<- pg_pharm+MY

map

```